Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2018-04-18 2017-Xudong-SMT/AVOCS-2018.tex
comments of Rev 2

In the long goal, we want to be able to check the equivalence between
open systems encoded as pNets. The equivalence between pNets is
"FH-bisimulation" taking the 
predicate of the open transitions into account each time matching such
open transitions. We foresee that the interplay with the SMT solver
that we use here for satisfiability of open transitions will be
similar with what we need to prove (symbolic) equivalence between open
transitions.

@@

\paragraph{Structure.}
In section
\ref{section:pnets} we give a description and a formal definition of
the pNet model, as found in previous publications. Then in section
\ref{section:examples} we present our use-case, based on a BIP
architecture from the nano-satellite case-study. 
Section \ref{section:op-semantics} recalls briefly the operational semantics
of pNet.
Section \ref{section:implementation} explains in details the algorithm
used to compute this semantics, including the interaction with Z3, and
shows the full result of the semantic computation on the running example.
Finally we conclude and discuss perspectives.
\ref{section:conclusion}.

@@

A pLTS is a labelled transition system with variables, which can be
manipulated, defined, or accessed inside states, actions, guards, and
assignments. 
%
%Without loss of generality and to simplify the formalization, we suppose 
%here that variables are local to 
%state: each state has its set of variables disjoint from the others. Transmitting 
%variable values from one state to the other can be done by explicit assignment. 
%
Each state has its set of variables called \emph{state variables}, 
which can only be modified by the assignment in transitions targeting this state. 
A global state variable of a pLTS is a variable $x$ satisfying
$\forall s\!\in\! S .\, x\!\in\! \vars(s)$. 
%
%Similarly, to simplify the management of variables and without loss of expressivity, we 
%suppose that transitions looping to the same state does not do assignments.
Note that we make no assumptions on finiteness of the set of states, nor
on finite branching of the transition relation.

@@

  \begin{definition}[pNets]\label{def-pnets}
    A pNet is a hierarchical structure where leaves are pLTSs and holes:
   %%  \footnote{Typing can 
%% be added to pNets by defining and checking their interfaces, called 
%% sort~\cite{HMZ-FORTE2016}}
$\pNet\triangleq \pLTS~|~\mylangle \pNet_i^{i\in I}, J, \symb{SV}_k^{k\in 
K}\myrangle$
where
\begin{itemize}
\item[$\bullet$] $\pNet_i^{i\in I}$ is the family of sub-pNets;
%  $\pNet_i^{i\in I}$ is a family of sub-pNets where $I\in\I_\P$ is the set over which 
%sub-pNets are indexed.

@@

\begin{definition}[Open transitions]
	\label{def:OpenTransitions}
	An \emph{open transition} over a set  $\langle S_i,s_{0 i}, \rightarrow_i\rangle^{i\in
	I}$ of LTSs, a
	set $J$ of holes with sorts $Sort_j^{j\in J}$, and a set of states $\mathcal{S}$ is a structure of the form:	
	\begin{mathpar}
	\inferrule*[myfraction=\reddottedrule]
	{\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
		\{\xrightarrow{b_j}_j\}^{j\in J}, \Pred\,, \Post}
	{s \xrightarrow {v}s'}\,,
	\end{mathpar}
	where $s, s'\in\mathcal{S}$ and, for all
        $i\in I$, $s_i{\xrightarrow{a_i}}_i s_i^{\prime}$ is a transition of the
	LTS $\langle S_i,s_{0 i}, \rightarrow_i\rangle$, and $\xrightarrow{b_j}_j$
        is a transition of the hole $j$, for any action $b_j$ in the
        sort $Sort_j$. \Pred\ is a predicate
	over the variables of the terms, labels, and states $s_i$,
        $b_j$, $s$, $v$. \Post\ is a set of equations that  
	hold \emph{after the open transition}, represented as a
        substitution $\{x_k\gets e_k\}^{k\in K}$  
	where $x_k$ are variables of $s'$ and $s'_i$, whereas $e_k$ are
        expressions over the other variables of the open transition.
\end{definition}

@@

\paragraph{Predicates.}
Let
$\mylangle\overline{\pNet},\overline{S},\symb{SV}_k^{k\in K} \myrangle$
be a pNet. Consider a synchronization vector $SV_k$, for $k\in K$. We build a
predicate $\MkPred$ relating
the actions of the involved sub-pNets and the resulting actions. This predicate verifies:
\begin{multline*}
  \MkPred(SV_k, a_i^{i\in I}, b_j^{j\in J}, v)\Longleftrightarrow
  %% \begin{array}{l}
  \exists {(a'_i)}^{i\in I},
          {(b'_j)}^{j\in J},v'.\\
          SV_k={(a'_i)}^{i\in I}, {(b'_j)}^{j\in J}\rightarrow v'
          \land
          \forall i\in I.\, a_i=a'_i\land \forall j \in J.\, b_j=b'_j \land v=v'
\end{multline*}

@@

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U\setminus E$; add $S$ in $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OT} = \mathit{MakeTransitions}(P, S)$;
        \State Store $\mathit{OT}$ in $L$; 
	\For {each $\mathit{OT} \in L$}
        \State Let $T$ be the target global state of $\mathit{OT}$
	\IfThen{($T \not\in U$ \&\& $T \not\in E$)}{Add $T$ into $U$;}
	\EndFor
\EndWhile
\State Prune the unsatisfiable open transitions in $L$ using the SMT solver;
\State \Return $\mathit{OA}=\mathit{Reachable}(\mathit{InitState}(P),L)$;

@@

At the root of the pNet, the predicate of each OT is translated
into SMTlib assertions, and checked for satisfiability.
%
The final open automaton is computed as the set of global states,
reachable after pruning the unsatisfiable OTs.
%
%% Unsatisfiable
%% OTs are pruned, and the final open automaton is computed as the set of
%% global states reachable by satisfiable OTs.

>>>>

In the long goal, we want to be able to check the equivalence between
open systems encoded as pNets. The equivalence between pNets is
"FH-bisimulation" \cite{henrio:Forte2016}, a dedicated version of
symbolic bisimulation taking the  
predicate of the open transitions into account when matching such
open transitions. We foresee that the interplay with the SMT solver
that we use here for satisfiability of open transitions will be
similar with what we need to prove (symbolic) equivalence between open
transitions.

@@

\paragraph{Structure.}
In section
\ref{section:pnets} we give a description and a formal definition of
the pNet model, as found in previous publications. Then in section
\ref{section:examples} we present our use-case, based on a BIP
architecture from the nano-satellite case-study. 
Section \ref{section:op-semantics} recalls briefly the operational semantics
of pNet.
Section \ref{section:implementation} explains in details the algorithm
used to compute this semantics, including the interaction with Z3, and
shows the full result of the semantic computation on the running example.
Finally we conclude and discuss perspectives in Section
\ref{section:conclusion}.

@@

A pLTS is a labelled transition system with variables, which can be
manipulated, defined, or accessed inside states, actions, guards, and
assignments. 
%
%Without loss of generality and to simplify the formalization, we suppose 
%here that variables are local to 
%state: each state has its set of variables disjoint from the others. Transmitting 
%variable values from one state to the other can be done by explicit assignment. 
%
Each state has its set of variables called \emph{state variables}, 
which can only be modified by the assignment in transitions targeting this state. 
A global state variable of a pLTS is a state variable defined in all states.
% $x$ satisfying $\forall s\!\in\! S .\, x\!\in\! \vars(s)$. 
%
%Similarly, to simplify the management of variables and without loss of expressivity, we 
%suppose that transitions looping to the same state does not do assignments.
Note that we make no assumptions on finiteness of the set of states, nor
on finite branching of the transition relation.

@@

  \begin{definition}[pNets]\label{def-pnets}
    A pNet is a hierarchical structure where leaves are pLTSs and holes:
   %%  \footnote{Typing can 
%% be added to pNets by defining and checking their interfaces, called 
%% sort~\cite{HMZ-FORTE2016}}
$\pNet\triangleq \pLTS~|~\mylangle \pNet_i^{i\in I}, J, \symb{SV}_k^{k\in 
K}\myrangle$, with $I, J, K$ potentialy infinite,
where
\begin{itemize}
\item[$\bullet$] $\pNet_i^{i\in I}$ is the family of sub-pNets;
%  $\pNet_i^{i\in I}$ is a family of sub-pNets where $I\in\I_\P$ is the set over which 
%sub-pNets are indexed.

@@

\begin{definition}[Open transitions]
	\label{def:OpenTransitions}
	An \emph{open transition} $OT$ over a set  $\langle S_i,s_{0 i}, \rightarrow_i\rangle^{i\in
	I}$ of LTSs, a
	set $J$ of holes, %with sorts $Sort_j^{j\in J}$
        and a set of states $\mathcal{S}$ is a structure of the form:	
	\begin{mathpar}
	\inferrule*[myfraction=\reddottedrule]
	{\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
		\{\xrightarrow{b_j}_j\}^{j\in J}, \Pred\,, \Post}
	{s \xrightarrow {v}s'}\,,
	\end{mathpar}
	where $s, s'\in\mathcal{S}$, the $a_i, b_j, \alpha$ are
        action expressions, $s_i{\xrightarrow{a_i}}_i s_i^{\prime}$ is
        a transition of the 
	LTS $\langle S_i,s_{0 i}, \rightarrow_i\rangle$, $b_j$
        is an action of the hole $j$, and $\alpha$ is the resulting action
        of $OT$. \Pred\ is a predicate
	over the variables of the terms, labels, and states $s_i$,
        $b_j$, $s$, $\alpha$. \Post\ is a set of equations that  
	hold \emph{after the open transition}, represented as a
        substitution $\{x_k\gets e_k\}^{k\in K}$  
	where $x_k$ are variables of $s'$ and $s'_i$, whereas $e_k$ are
        expressions over the other variables of the open transition.
\end{definition}

@@

\paragraph{Predicates.}
Let
$\mylangle\overline{\pNet},J,\symb{SV}_k^{k\in K} \myrangle$
be a pNet. Consider a synchronization vector $SV_k$, for $k\in K$. We build a
predicate $\MkPred$ relating
the actions of the involved sub-pNets and the resulting actions. This predicate verifies:
\begin{multline*}
  \MkPred(SV_k, a_i^{i\in I}, b_j^{j\in J}, v)\Longleftrightarrow
  %% \begin{array}{l}
  \exists {(a'_i)}^{i\in I},
          {(b'_j)}^{j\in J},v'.\\
          SV_k={(a'_i)}^{i\in I}, {(b'_j)}^{j\in J}\rightarrow v'
          \land
          \forall i\in I.\, a_i=a'_i\land \forall j \in J.\, b_j=b'_j \land v=v'
\end{multline*}

@@

In \cite{henrio:Forte2016} we also
proved that a pNet with finitely many pLTSs, holes, and
synchronisation vectors, and with finite pLTSs, has a finite
semantics. But all the elements of such pNets and open automata being
symbolic, they can represent many classes of unbounded systems.

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
%%        \State Store $\mathit{OT}$ in $L$; 
	\For {each $\mathit{OT} \in \mathit{OTs}$}
        Check satisfiability of $\mathit{OT}$ using the SMT solver;
        \IfThen {$SAT(\mathit{OT})$}
                {\State \{Add $\mathit{OT}$ to $L$;
                  \State Let $S'$ be the target global state of $\mathit{OT}$
	          \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
                  \}}
	\EndFor
\EndWhile
%%\State Prune the unsatisfiable open transitions in $L$ using the SMT solver;
\State \Return $\mathit{OA}=(\mathit{InitState}(P),L)$;

@@

At the root of the pNet, the predicate of each OT is translated
into SMTlib assertions, and checked for satisfiability.
%
The final open automaton includes all satisfiable OTs, and the set of
reachable global states.

%
%% Unsatisfiable
%% OTs are pruned, and the final open automaton is computed as the set of
%% global states reachable by satisfiable OTs.
